---
title: 'Assignment 11 - Detecting Overfitting and Overfitting Issues'
author: 'Sabah Khan'
date: "Due: April 14, 2026"
output:
  html_document:           # output document format
    toc: yes               # add table contents
    toc_float: yes         # toc_property: floating
    toc_depth: 4           # depth of TOC headings
    fig_width: 6           # global figure width
    fig_height: 4          # global figure height
    fig_caption: yes       # add figure caption
    number_sections: yes   # numbering section headings
    toc_collapsed: yes     # TOC subheading clapsing
    code_folding: hide     # folding/showing code 
    code_download: yes     # allow to download complete RMarkdown source code
    smooth_scroll: yes     # scrolling text of the document
    theme: lumen           # visual theme for HTML document only
    highlight: tango       # code syntax hightlighting styles
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
  word_document:
    toc: yes
    toc_depth: '4'
---

```{css, echo = FALSE}
div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 24px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";
}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.

if (!require("knitr")) {                      # use conditional statement to detect
   install.packages("knitr")                  # whether a package was installed in
   library(knitr)     
if (!require("imager"))
  installed.packages("imager")
  library(imager)
  if (!require("latex2exp"))
  installed.packages("latex2exp")
  library(latex2exp)
                                                # your machine. If not, install it and
}                                             # load it to the working directory.
#
knitr::opts_chunk$set(echo = TRUE,            # include code chunk in the output file
                      warning = FALSE,        # sometimes, you code may produce warning messages,
                                              # you can choose to include the warning messages in
                                              # the output file. 
                      results = TRUE,         # you can also decide whether to include the output
                                              # in the output file.
                      message = FALSE,        # suppress messages 
                      comment = NA            # remove the default leading hash tags in the output
                      )   
```


# Part a

```{r}

time2event <- c(12.4, 18.7, 25.3, 30.1, 33.5, 35.2, 38.9, 40.3, 42.7, 45.1, 47.6, 49.8, 52.4, 55.0, 
57.3, 60.2, 62.8, 65.1, 67.9, 70.5, 72.3, 75.6, 78.2, 80.9, 83.4, 85.7, 88.1, 90.6, 
93.2, 95.8, 98.4, 101.0, 104.5, 107.3, 110.6, 113.2, 116.8, 120.1, 123.7, 127.4,
130.9, 134.5, 138.2, 142.0, 146.3, 150.7, 155.2, 160.8, 168.4, 175.9)

df <- data.frame(values = time2event)
# print(df) print if necessary. Not printing for html output.data frame created from vector.

colnames(df) <- "time"


x <- df$time



# Log-likelihood for Weibull distribution
# lambda = scale, beta = shape
log_likelihood_weibull <- function(param) {
  lambda <- param[1]  # scale
  beta   <- param[2]  # shape
  
  # Input validation
  if (lambda <= 0 || beta <= 0) stop("Parameters must be positive.")
  if (any(x <= 0)) stop("All observations must be positive.")
  
  n <- length(x)
  sum_logx <- sum(log(x))
  sum_x_beta <- sum((x / lambda) ^ beta)
  
  ll <- n * log(beta) -
        n * beta * log(lambda) +
        (beta - 1) * sum_logx -
        sum_x_beta
  
  return(ll)
}

# Score equations for Weibull distribution
# lambda = scale, beta = shape
score_equ_weibull <- function(par) {
  lambda <- par[1]  # scale
  beta   <- par[2]  # shape
  
  if (lambda <= 0 || beta <= 0) stop("Parameters must be positive.")
  if (any(x <= 0)) stop("All observations must be positive.")
  
  n <- length(x)
  log_x <- log(x)
  log_x_over_lambda <- log_x - log(lambda)  # precompute log(x / lambda)
  x_over_lambda <- x / lambda
  x_over_lambda_beta <- x_over_lambda ^ beta
  
  # Score for beta (shape)
  score_beta <- (n / beta) -
                n * log(lambda) +
                sum(log_x) -
                sum(x_over_lambda_beta * log_x_over_lambda)
  
  # Score for lambda (scale)
  score_lambda <- (beta / lambda) * (sum(x_over_lambda_beta) - n)
  
  return(c(score_lambda, score_beta))
}

# Fisher Information Matrix for Weibull (lambda = scale, beta = shape)
FisherInfo_weibull <- function(par) {
  lambda <- par[1]  # scale
  beta   <- par[2]  # shape
  
  if (lambda <= 0 || beta <= 0) stop("Parameters must be positive.")
  if (any(x <= 0)) stop("All observations must be positive.")
  
  n <- length(x)
  log_x <- log(x)
  log_x_over_lambda <- log_x - log(lambda)
  x_over_lambda <- x / lambda
  x_over_lambda_beta <- x_over_lambda ^ beta
  
  # Second derivatives (observed Hessian components)
  d2_ll_lambda2 <- -n * beta / (lambda^2) - beta * (beta + 1) / (lambda^2) * sum(x_over_lambda_beta)
  d2_ll_beta2   <- -n / (beta^2) - sum(x_over_lambda_beta * (log_x_over_lambda ^ 2))
  d2_ll_lambdabeta <- -n / lambda + (1 / lambda) * sum(x_over_lambda_beta * 
                                                         (1 + beta * log_x_over_lambda))
  
  # Observed Fisher Information = - Hessian
  I_11 <- -d2_ll_lambda2
  I_22 <- -d2_ll_beta2
  I_12 <- -d2_ll_lambdabeta
  
  Fisher <- matrix(c(I_11, I_12, I_12, I_22), nrow = 2, byrow = TRUE)
  return(Fisher)
}

#Maximum Likelihood Estimate using optim

# Weibull MLE with lambda = scale, beta = shape

# Data
x <- df$time
n <- length(x)


 # initial values
###
 
 # Initial values 
# lambda_init: scale guess (mean)
# beta_init: shape guess (1.0 corresponds to exponential)
lambda_init <- mean(x)
beta_init   <- 1.0 

# Run optim for Weibull MLE
result <- optim(
  par = c(lambda_init, beta_init),   # initial guesses: (lambda, beta)
  fn = log_likelihood_weibull,       # log-likelihood function
  gr = score_equ_weibull,            # score function (gradient)
  method = "BFGS",                    # quasi-Newton method
  hessian = TRUE,                     # return Hessian matrix
  control = list(
    fnscale = -1,    # maximize instead of minimize
    maxit = 1000,    # stopping rule 1: max iterations
    reltol = 1e-8,   # stopping rule 2: precision control
    REPORT = 10      # print progress every 10 iterations
  )
)

# View results
result
 
Fisher <- FisherInfo_weibull(result$par)
Fisher

# Which is consistent with the hessian matrix in the output of optim()

varcov <- solve(-result$hessian)   # solve() is a matrix inversion function
varcov

#The variance and covariance matrix can be approximated by inverting the Fisher Information Matrix or the inverse of the negative #Hessian matrix returned from optim()

MLE <- result$par
MLE

```
The MLE's for λ and β are 99.01 and 2.21


# Parts b, c, and d

```{r}

# Log-likelihood for Weibull with beta = 1 (Exponential case)
loglik_weibull_beta1 <- function(lambda) {
  if (lambda <= 0) stop("Scale parameter lambda must be positive.")
  ll <- -n * log(lambda) - sum(x) / lambda
  return(ll)
}

# Score function (gradient) for Weibull with beta = 1
score_weibull_beta1 <- function(lambda) {
  if (lambda <= 0) stop("Scale parameter lambda must be positive.")
  score <- -n / lambda + sum(x) / (lambda^2)
  return(score)
}

#data
x <- df$time
n <- length(x)

# Optimization
result0 <- optim(
  par = mean(x),                     # initial guess for lambda
  fn = loglik_weibull_beta1,          # log-likelihood function
  gr = score_weibull_beta1,           # score function
  method = "BFGS",                    # quasi-Newton method
  hessian = TRUE,                     # return Hessian
  control = list(
    fnscale = -1,                     # maximize instead of minimize
    maxit = 1000,                      # max iterations
    reltol = 1e-8,                     # convergence tolerance
    REPORT = 10                        # print progress every 10 iterations
  )
)

result0

```
The restrictive MLE for lambda (when beta = 1 and we are dealing with an exponential distribution) is 87.61



 
```{r}

x <- df$time
 n = length(x)
 ##
 LR = -2*(loglik_weibull_beta1(result0$par) - log_likelihood_weibull(result$par))
 LR
 
 # The above is the likelihood ratio statistic
 
 pval <- 1 - pchisq(LR, df = 1)
pval
 
```

The p value is tiny, "4.373534e-09," therefore we reject the null hypothesis that beta = "1". 
 
```{r}
# Wald test
alpha.hat  <- result$par[1]     # based on the unrestricted likelihood estimation
se.alpha <- sqrt(varcov[1, 1])  # the square root of the top-left diagonal value
W <- (alpha.hat - 1)^2/(se.alpha)^2   # Wald test statistic
##
pval.W <- 1 - pchisq(W, df = 1)
c(W = W, pval.W = pval.W)

```

The p value for the Wald test is also zero, implying the rejection of the null hypothesis.

# Part e

The two tests generated the same results - namely, that the recommended model for the data is a Weibull model, whose scale parameter is not fixed at 1 (and is therefore not exponentially distributed data), and whose maximum likelihood estimate for scale parameter is 99.01 and whose maximum likelihood estimate for scale parameter is 2.21.

Below is a density curve for the data. Time to failure has a right skewed distribution, with the highest likelihood of failure occurring at between 70 and 90 hours

```{r}
x <- df$time

result <- optim(
  par = c(mean(x), 1.0),             # initial guesses
  fn = log_likelihood_weibull,       # log-likelihood function
  gr = score_equ_weibull,            # score function
  method = "BFGS",
  hessian = TRUE,
  control = list(fnscale = -1)
)

lambda_hat <- result$par[1]  # MLE scale
beta_hat   <- result$par[2]  # MLE shape

# Histogram of data
hist(x,
     breaks = 14,
     probability = TRUE,  # scale to density
     xlab = "Observed times",
     main = "Weibull Fit to Data",
     col = "lightgray",
     border = "white")

# Add Weibull density curve
curve(dweibull(x, shape = beta_hat, scale = lambda_hat),
      from = min(x), to = max(x),
      col = "red", lwd = 2, add = TRUE)

# Add kernel density for comparison
lines(density(x), col = "blue", lwd = 2, lty = 2)

# Legend
legend("topright",
       legend = c("Weibull MLE", "Kernel Density"),
       col = c("red", "blue"),
       lwd = 2, lty = c(1, 2))

```