---
title: 'Assignment 8 - Bootstrap Confidence Intervals '
author: 'Sabah Khan'
date: "Due: March 31, 2026"
output:
  html_document:           # output document format
    toc: yes               # add table contents
    toc_float: yes         # toc_property: floating
    toc_depth: 4           # depth of TOC headings
    fig_width: 6           # global figure width
    fig_height: 4          # global figure height
    fig_caption: yes       # add figure caption
    number_sections: yes   # numbering section headings
    toc_collapsed: yes     # TOC subheading clapsing
    code_folding: hide     # folding/showing code 
    code_download: yes     # allow to download complete RMarkdown source code
    smooth_scroll: yes     # scrolling text of the document
    theme: lumen           # visual theme for HTML document only
    highlight: tango       # code syntax hightlighting styles
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
  word_document:
    toc: yes
    toc_depth: '4'
---

```{css, echo = FALSE}
div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 24px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";
}
```

Part a

```{r setup}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.

if (!require("knitr")) {                      # use conditional statement to detect
   install.packages("knitr")                  # whether a package was installed in
   library(knitr)     
if (!require("imager"))
  installed.packages("imager")
  library(imager)
 if (!require("boot")) 
  library(boot)
  
                                                # your machine. If not, install it and
}                                             # load it to the working directory.
#
knitr::opts_chunk$set(echo = TRUE,            # include code chunk in the output file
                      warning = TRUE,        # sometimes, you code may produce warning messages,
                                              # you can choose to include the warning messages in
                                              # the output file. 
                      results = TRUE,         # you can also decide whether to include the output
                                              # in the output file.
                      message = TRUE,        # suppress messages 
                      comment = NA            # remove the default leading hash tags in the output
                      )   

#I cannot seem to supress the outputted warning messages. I have tried all combinations of warning = and message = in the above code. 

# data 
data_values <- c(0.85, 1.23, 0.92, 3.45, 2.11, 1.56, 4.89, 2.34, 1.78, 6.72, 0.95, 1.34, 8.91, 
2.67, 1.89, 5.43, 1.12, 3.78, 2.45, 7.65, 1.05, 1.45, 12.34, 2.89, 2.01, 4.56, 
1.23, 4.32, 2.67, 9.87, 0.99, 1.56, 15.23, 3.12, 2.34, 3.89, 1.34, 5.67, 2.89, 
11.45, 1.12, 1.67, 18.90, 3.45, 2.56, 3.45, 1.45, 6.78, 3.12, 14.56, 1.23, 1.78, 
22.34, 3.78, 2.78)


# Negative log-likelihood function for lognormal
neg_log_likelihood <- function(params, x) {
  meanlog <- params[1]
  sdlog   <- params[2]
  
  # no return of negative values
  if (sdlog <= 0) return(Inf)
  
  # Compute negative log-likelihood
  -sum(dlnorm(x, meanlog = meanlog, sdlog = sdlog, log = TRUE))
}

# Initial guesses for meanlog and sdlog
init_params <- c(mean(log(data_values)), sd(log(data_values)))

# Run optim
fit <- optim(
  par = init_params,
  fn = neg_log_likelihood,
  x = data_values,
  method = "L-BFGS-B",          # Bounded optimization
  lower = c(-Inf, 1e-8),        # sdlog > 0
  control = list(fnscale = 1)   # Minimize
)

# Check convergence
if (fit$convergence != 0) {
  warning("Optimization may not have converged.")
}

# Extract fitted parameters
fitted_meanlog <- fit$par[1]
fitted_sdlog   <- fit$par[2]

cat("Fitted meanlog:", fitted_meanlog, "\n")
cat("Fitted sdlog:", fitted_sdlog, "\n")


# Manual implementation
B <- 5000
n <- length(data_values)
boot_ln_means <- numeric(B)     # empty vector to store bootstrap lognormal means

for(b in 1:B) {
  boot_sample <- sample(data_values, n, replace=TRUE)  # generating bootstrap sample
  boot_ln_means[b] <- mean(boot_sample)            # calculate bootstrap means
}

theta_hat <- mean(log(data_values))   # same lognormal mean from the original sample
se_boot <- sd(boot_ln_means)
z <- qnorm(0.975)
ci_normal <- c(LCI = theta_hat - z*se_boot, xbar = theta_hat, UCL=theta_hat + z*se_boot)
ci_normal

###
hist(boot_ln_means,                     # data used for histogram
     breaks = 14,                            # specify number of vertical bars
     probability = TRUE,
       xlab = "Bootstrap estimated sample lognormal means",      # change the label of x-axis
      # add a title to the histogram
        main="Bootstrap Sampling Distribution of Estimated Sample Lognormal Means",
          cex.main = 0.9,
       col.main = "navy")   
lines(density(boot_ln_means), col = "skyblue", lwd = 2)

```

Well, as expected, the bootstrapped estimates of the sample lognormal means show a normal distribution. This is defined in the assignment; that Y=ln(X)∼N(μ,σ2).


Part b

```{r}

library(boot)

# Define statistic function
mean_stat <- function(data_values, indices){
  sample_data <- data_values[indices]
  return(mean(log(sample_data)))
}


# Perform bootstrap
boot_result <- boot(data_values, mean_stat, R=5000)

# Percentile CI
boot.ci(boot_result, type="perc")

# Manual calculation
boot_vals <- boot_result$t
alpha <- 0.05
ci_perc <- quantile(boot_vals, probs=c(alpha/2, 1-alpha/2))
ci_perc


hist(boot_vals,                     # data used for histogram
     breaks = 14,                            # specify number of vertical bars
     probability = TRUE,
       xlab = "Bootstrap estimated sample lognormal means - percentile CI",      # change the label of x-axis
      # add a title to the histogram
        main="Bootstrap Sampling Distribution of Estimated Sample Lognormal Means - percentile CI",
          cex.main = 0.9,
       col.main = "navy")   
lines(density(boot_vals), col = "skyblue", lwd = 2)

```

Part c

```{r}

library(boot)

# Define statistic function
mean_stat_c <- function(data_values, indices){
  return(mean(log(data_values)))}

set.seed(123)
data <- dlnorm(55, meanlog = 1.08962, sdlog = 0.8363173)
boot_result_c <- boot(data_values, mean_stat_c, R=5000)


# BCa CI
bca_ci <- boot.ci(boot_result_c, type="bca")
bca_ci


```

Part d

The Central Limit Theorem (CLT) states that when the sample size is sufficiently large, the sampling distribution of the sample mean will be approximately normal, regardless of the shape of the population distribution, provided the population has a finite variance. This is true of the lognormal distribution anyway; the bootstrapped estimates of the sample lognormal means are normally distributed.



```{r}


# Normal Bootstrap CI for normal
set.seed(123)
example1.dat <- rnorm(55, mean=1.089672, sd=0.8363173)  # simulate a normal random sample using lognormal mean for mean and sdlog for sd

# Manual implementation
B <- 5000
n <- length(example1.dat)
boot_means_d <- numeric(B)     # empty vector to store bootstrap sample means

for(b in 1:B) {
  boot_sample_d <- sample(example1.dat, n, replace=TRUE)  # generating bootstrap sample
  boot_means_d[b] <- mean(boot_sample_d)            # calculate bootstrap means
}

theta_hat <- mean(example1.dat)   # same mean from the original sample
se_boot <- sd(boot_means_d)
z <- qnorm(0.975)
ci_normal <- c(LCI = theta_hat - z*se_boot, xbar = theta_hat, UCL=theta_hat + z*se_boot)
ci_normal


# Compare different methods
boot.ci(boot_result, type=c("norm", "perc", "bca"))

```
The normal CI constructed in part d is the widest, the bca is the narrowest (although I struggled with the coding above I was able to get output for comparison for bca CI's).

The symmetry looks best on the percentile CI histogram - just eyeballing it, but knowing that we want sensitivity to distributional skewness and stability, I would say the bca method is probably the most idea. 

