---
title: 'Assignment 9 - Hypothesis Testing and Power and Sample size Determination'
author: 'Sabah Khan'
date: "Due: April 7, 2026"
output:
  html_document:           # output document format
    toc: yes               # add table contents
    toc_float: yes         # toc_property: floating
    toc_depth: 4           # depth of TOC headings
    fig_width: 6           # global figure width
    fig_height: 4          # global figure height
    fig_caption: yes       # add figure caption
    number_sections: yes   # numbering section headings
    toc_collapsed: yes     # TOC subheading clapsing
    code_folding: hide     # folding/showing code 
    code_download: yes     # allow to download complete RMarkdown source code
    smooth_scroll: yes     # scrolling text of the document
    theme: lumen           # visual theme for HTML document only
    highlight: tango       # code syntax hightlighting styles
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
  word_document:
    toc: yes
    toc_depth: '4'
---

```{css, echo = FALSE}
div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 24px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";
}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.

if (!require("knitr")) {                      # use conditional statement to detect
   install.packages("knitr")                  # whether a package was installed in
   library(knitr)     
if (!require("imager"))
  installed.packages("imager")
  library(imager)
  if (!require("latex2exp"))
  installed.packages("latex2exp")
  library(latex2exp)
                                                # your machine. If not, install it and
}                                             # load it to the working directory.
#
knitr::opts_chunk$set(echo = TRUE,            # include code chunk in the output file
                      warning = FALSE,        # sometimes, you code may produce warning messages,
                                              # you can choose to include the warning messages in
                                              # the output file. 
                      results = TRUE,         # you can also decide whether to include the output
                                              # in the output file.
                      message = FALSE,        # suppress messages 
                      comment = NA            # remove the default leading hash tags in the output
                      )   
```


# Part a

Since we are only given the average cholesterol reduction value of 29 mg/dL for the current standard drug, but we don't have any information about a sample size or a study for the current standard drug, we can use a one sample t-test for the new drug, using u=25. 

n=5,x¯=29,s=15,μ=25, and α=0.05

Hypotheses:

H0:μ=25 (no effect)
H1:μ>25 (one-tailed, CholestFix reduces cholesterol by more than 25 mg/dL)

Test Statistic:

n=5, x¯=29, s=15, μ0=25

t=29−25/(15/√5) = 4/6.708 = 0.596

df=n−1=4

tα=0.05, df=4 = 2.132

Decision: since t= 0.596 < 2.132, fail to reject H0


Verification using R:

```{r}

# One-sample t-test in R
set.seed(123)
simulated_ct <- rnorm(5, mean = 29, sd = 15)
result <- t.test(simulated_ct, mu = 25, 
                 alternative = "greater")

# Output
print(result)

###
cat("\nManual t-value: 0.596",
    "\nR t-value:", round(result$statistic, 3),
    "\nCritical t-value (0.05): 2.132",
    "\nR p-value:", round(result$p.value, 4))

#I am not sure why the R t-value is so drastically different from my t value. I followed your code for the one sample t-test. However, the p value outputted by R is still  > 0.05, so the decision remains the same - fail to reject null. Indicating that there is not a significant difference between the standard drug and CholestFix. Although R is outputting a mean that is higher than 25, the 95% confidence interval for the estimated mean contains 20.305, so we cannot say CholestFix is different from the standard drug.

```
Although the R t-value is quite different from the manually calculated t-value, the p-value outputted by R is still  > 0.05, so the decision would be the same whether the calculation is done manually or in R (in this case at least) - fail to reject null. Indicating that there is not a significant difference between the standard drug and CholestFix. Although R is outputting a mean that is higher than 25, the 95% confidence interval for the estimated mean contains 20.305, so we cannot say CholestFix is different from the standard drug.

# Part b

Manually, we can say, H0:μ=μ0,H1:μ>μ0

n=50,α=0.05,σ=15 (one-sided)

λ=29−25/(15/√50)= 7.45 and δ=29−25= 4.

Using software to calculate power:

```{r}
# Power calculation for one-sample t-test
power_calc <- power.t.test(n = 50, 
                          delta = 4,    # true difference between the means in H0 and Ha
                          sd = 15,      
                          sig.level = 0.05,
                          type = "one.sample",
                          alternative = "one.sided")
print(power_calc)

```
The probability that we'd detect a true improvement is 0.585 or about 59%, given a sample size of n=50, indicating that our sample size in part a was far too small to detect a true improvement.

# Part c

```{r}
# Sample size calculation
sample_size <- power.t.test(power = 0.80, 
                            delta = 4,
                           sd = 15,
                           sig.level = 0.05,
                           type = "one.sample",
                           alternative = "one.sided")
cat("\nRequired sample size for 80% power:", ceiling(sample_size$n))

```
The required sample size to detect an effect size of 4 mg/dL with a power of 1−β=0.8 and a significance level of α=0.05 is n = 89, which makes sense because in part B we had a sample size of 50, which is less than 89, and so we were getting power less than 0.80 because our sample size was still not large enough. 

# Part d

```{r}
# given info
n = 5
power = 0.80
sig = 15
delta = 4
lambda = 0.596
sig.level = 0.05
##
crit = qt(0.975, n-1)
##
power4 = 1-pt(crit, df = 4, ncp = lambda) + pt(-crit, df = 4, ncp = lambda)
n =((qnorm(0.975)-qnorm(0.10))*sig/delta)^2 
data.frame(cbind(power = power, n = n))

```


```{r}
# given info
n = 50
power = 0.80
sig = 15
delta = 4
lambda = 1.886
sig.level = 0.05
##
crit = qt(0.975, n-1)
##
power50 = 1-pt(crit, df = 49, ncp = lambda) + pt(-crit, df = 49, ncp = lambda)
n =((qnorm(0.975)-qnorm(0.10))*sig/delta)^2 
data.frame(cbind(power = power, n = n))

#The sample size required here by manually inputting a power calculation is 148. Again, this differs from the sample size calculated in part c, and I'm not sure why. The only thing I can think of is that in Part C I have specified that it is a one sided test, and here I am not sure how to specify this, or if it even needs to be specified. 

```
```{r}

#I could not find anything in the lecture note about plotting a power curve. However, I tried my best. The argument for alternative in the code below requires that I specify two sided, so I'm not sure this is the type of graphic you are looking for. 

# Install pwr package if not already installed
if (!require(pwr)) {
  install.packages("pwr")
  library(pwr)
}

# Parameters
sigma <- 15          # Known standard deviation
mu0 <- 25           # Null hypothesis mean
mu1 <- 29           # Alternative hypothesis mean
alpha <- 0.05        # Significance level
n_values <- seq(5, 50, by = 5)  # Sample sizes to evaluate

# Calculate effect size (Cohen's d)
effect_size <- (mu1 - mu0) / sigma

# Compute power for each sample size
power_values <- sapply(n_values, function(n) {
  pwr.t.test(n = n, d = effect_size, sig.level = alpha, type = "one.sample", alternative = "two.sided")$power
})

# Plot the power curve
plot(n_values, power_values, type = "b", pch = 19, col = "blue",
     xlab = "Sample Size (n)", ylab = "Statistical Power",
     main = "Power Curve for One-Sample t-Test")
abline(h = 0.8, col = "red", lty = 2)  # Common 80% power threshold
grid()
```
