---
title: 'Assignment 12: Bootstrap Likelihood Ratio Test (BLRT)'
author: 'Sabah Khan'
date: "Due: April 21, 2026"
output:
  html_document:           # output document format
    toc: yes               # add table contents
    toc_float: yes         # toc_property: floating
    toc_depth: 4           # depth of TOC headings
    fig_width: 6           # global figure width
    fig_height: 4          # global figure height
    fig_caption: yes       # add figure caption
    number_sections: yes   # numbering section headings
    toc_collapsed: yes     # TOC subheading clapsing
    code_folding: hide     # folding/showing code 
    code_download: yes     # allow to download complete RMarkdown source code
    smooth_scroll: yes     # scrolling text of the document
    theme: lumen           # visual theme for HTML document only
    highlight: tango       # code syntax hightlighting styles
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
  word_document:
    toc: yes
    toc_depth: '4'
---

```{css, echo = FALSE}
div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 24px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";
}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.

if (!require("knitr")) {                      # use conditional statement to detect
   install.packages("knitr")                  # whether a package was installed in
   library(knitr)     
if (!require("imager"))
  installed.packages("imager")
  library(imager)
  if (!require("latex2exp"))
  installed.packages("latex2exp")
  library(latex2exp)
                                                # your machine. If not, install it and
}                                             # load it to the working directory.
#
knitr::opts_chunk$set(echo = TRUE,            # include code chunk in the output file
                      warning = FALSE,        # sometimes, you code may produce warning messages,
                                              # you can choose to include the warning messages in
                                              # the output file. 
                      results = TRUE,         # you can also decide whether to include the output
                                              # in the output file.
                      message = FALSE,        # suppress messages 
                      comment = NA            # remove the default leading hash tags in the output
                      )   
```


# Part a

```{r}

time2event <- c(5.2,  7.8,  9.1, 11.3, 12.5, 13.0, 14.2, 15.1, 15.9, 16.7, 17.2, 17.8, 18.4, 18.9, 
  19.3, 19.7, 20.2, 20.6, 21.0, 21.5, 21.9, 22.3, 22.7, 23.1, 23.5, 23.9, 24.3, 24.7, 
  25.1, 25.5, 25.9, 26.3, 26.7, 27.1, 27.5, 27.9, 28.3, 28.7, 29.1, 29.5, 29.9, 30.3, 
  30.7, 31.1, 31.5, 31.9, 32.3, 32.7, 33.1, 33.5, 33.9, 34.3, 34.7, 35.1, 35.5, 35.9, 
  36.3, 36.7, 37.1, 37.5, 37.9, 38.3, 38.7, 39.1, 39.5, 39.9, 40.3, 40.7, 41.1, 41.5,
  41.9, 42.3, 42.7, 43.1, 43.5)

df <- data.frame(values = time2event)
# print(df) print if necessary. Not printing for html output.data frame created from vector.

colnames(df) <- "time"


x <- df$time



# Log-likelihood for Weibull distribution
# lambda = scale, beta = shape
log_likelihood_weibull <- function(param) {
  lambda <- param[1]  # scale
  beta   <- param[2]  # shape
  
  # Input validation
  if (lambda <= 0 || beta <= 0) stop("Parameters must be positive.")
  if (any(x <= 0)) stop("All observations must be positive.")
  
  n <- length(x)
  sum_logx <- sum(log(x))
  sum_x_beta <- sum((x / lambda) ^ beta)
  
  ll <- n * log(beta) -
        n * beta * log(lambda) +
        (beta - 1) * sum_logx -
        sum_x_beta
  
  return(ll)
}

# Score equations for Weibull distribution
# lambda = scale, beta = shape
score_equ_weibull <- function(par) {
  lambda <- par[1]  # scale
  beta   <- par[2]  # shape
  
  if (lambda <= 0 || beta <= 0) stop("Parameters must be positive.")
  if (any(x <= 0)) stop("All observations must be positive.")
  
  n <- length(x)
  log_x <- log(x)
  log_x_over_lambda <- log_x - log(lambda)  # precompute log(x / lambda)
  x_over_lambda <- x / lambda
  x_over_lambda_beta <- x_over_lambda ^ beta
  
  # Score for beta (shape)
  score_beta <- (n / beta) -
                n * log(lambda) +
                sum(log_x) -
                sum(x_over_lambda_beta * log_x_over_lambda)
  
  # Score for lambda (scale)
  score_lambda <- (beta / lambda) * (sum(x_over_lambda_beta) - n)
  
  return(c(score_lambda, score_beta))
}

# Fisher Information Matrix for Weibull (lambda = scale, beta = shape)
FisherInfo_weibull <- function(par) {
  lambda <- par[1]  # scale
  beta   <- par[2]  # shape
  
  if (lambda <= 0 || beta <= 0) stop("Parameters must be positive.")
  if (any(x <= 0)) stop("All observations must be positive.")
  
  n <- length(x)
  log_x <- log(x)
  log_x_over_lambda <- log_x - log(lambda)
  x_over_lambda <- x / lambda
  x_over_lambda_beta <- x_over_lambda ^ beta
  
  # Second derivatives (observed Hessian components)
  d2_ll_lambda2 <- -n * beta / (lambda^2) - beta * (beta + 1) / (lambda^2) * sum(x_over_lambda_beta)
  d2_ll_beta2   <- -n / (beta^2) - sum(x_over_lambda_beta * (log_x_over_lambda ^ 2))
  d2_ll_lambdabeta <- -n / lambda + (1 / lambda) * sum(x_over_lambda_beta * 
                                                         (1 + beta * log_x_over_lambda))
  
  # Observed Fisher Information = - Hessian
  I_11 <- -d2_ll_lambda2
  I_22 <- -d2_ll_beta2
  I_12 <- -d2_ll_lambdabeta
  
  Fisher <- matrix(c(I_11, I_12, I_12, I_22), nrow = 2, byrow = TRUE)
  return(Fisher)
}

#Maximum Likelihood Estimate using optim

# Weibull MLE with lambda = scale, beta = shape

# Data
x <- df$time
n <- length(x)


 # initial values
###
 
 # Initial values 
# lambda_init: scale guess (mean)
# beta_init: shape guess (1.0 corresponds to exponential)
lambda_init <- mean(x)
beta_init   <- 1.0 

# Run optim for Weibull MLE
result <- optim(
  par = c(lambda_init, beta_init),   # initial guesses: (lambda, beta)
  fn = log_likelihood_weibull,       # log-likelihood function
  gr = score_equ_weibull,            # score function (gradient)
  method = "BFGS",                    # quasi-Newton method
  hessian = TRUE,                     # return Hessian matrix
  control = list(
    fnscale = -1,    # maximize instead of minimize
    maxit = 1000,    # stopping rule 1: max iterations
    reltol = 1e-8,   # stopping rule 2: precision control
    REPORT = 10      # print progress every 10 iterations
  )
)

# View results
result
 
Fisher <- FisherInfo_weibull(result$par)
Fisher

# Which is consistent with the hessian matrix in the output of optim()

varcov <- solve(-result$hessian)   # solve() is a matrix inversion function
varcov

#The variance and covariance matrix can be approximated by inverting the Fisher Information Matrix or the inverse of the negative #Hessian matrix returned from optim()

MLE <- result$par
MLE

```
The MLE's for λ and β are 31.42 and 3.37


# Parts b, c, and d

```{r}

# Log-likelihood for Weibull with beta = 1 (Exponential case)
loglik_weibull_beta1 <- function(lambda) {
  if (lambda <= 0) stop("Scale parameter lambda must be positive.")
  ll <- -n * log(lambda) - sum(x) / lambda
  return(ll)
}

# Score function (gradient) for Weibull with beta = 1
score_weibull_beta1 <- function(lambda) {
  if (lambda <= 0) stop("Scale parameter lambda must be positive.")
  score <- -n / lambda + sum(x) / (lambda^2)
  return(score)
}

#data
x <- df$time
n <- length(x)

# Optimization
result0 <- optim(
  par = mean(x),                     # initial guess for lambda
  fn = loglik_weibull_beta1,          # log-likelihood function
  gr = score_weibull_beta1,           # score function
  method = "BFGS",                    # quasi-Newton method
  hessian = TRUE,                     # return Hessian
  control = list(
    fnscale = -1,                     # maximize instead of minimize
    maxit = 1000,                      # max iterations
    reltol = 1e-8,                     # convergence tolerance
    REPORT = 10                        # print progress every 10 iterations
  )
)

result0

```
The restrictive MLE for lambda (when beta = 1 and we are dealing with an exponential distribution) is 28.185



 
```{r}

x <- df$time
 n = length(x)
 ##
 LR = -2*(loglik_weibull_beta1(result0$par) - log_likelihood_weibull(result$par))
 LR
 
 # The above is the likelihood ratio statistic
 
 pval <- 1 - pchisq(LR, df = 1)
pval
 
```

The p value is 0, therefore we reject the null hypothesis that beta = "1". 
 
```{r}
# BLRT 

bootstrap_lrt_weibull <- function(x, B = 9999,
                                  beta_start = 3.37,
                                  lambda_start = 31.42) {
    n <- length(x)
    
    # Log-likelihoods 
    
    loglik_weibull <- function(par, x) {
        beta <- par[1]
        lambda <- par[2]
        
        if (beta <= 0 || lambda <= 0) return(-Inf)
        
        sum(dweibull(x, shape = beta, scale = lambda, log = TRUE))
    }
    
    loglik_exp <- function(lambda, x) {
        if (lambda <= 0) return(-Inf)
        sum(dexp(x, rate = 1/lambda, log = TRUE))
    }
    
    # Fit to original data 
    
    fit_alt <- optim(
        par = c(beta_start, lambda_start),
        fn = function(p) -loglik_weibull(p, x),
        method = "L-BFGS-B",
        lower = c(1e-6, 1e-6)
    )
    
    beta_hat <- fit_alt$par[1]
    lambda_hat <- fit_alt$par[2]
    logL_hat <- -fit_alt$value
    
    # Exponential MLE
    lambda0_hat <- mean(x)
    logL0 <- loglik_exp(lambda0_hat, x)
    
    # Observed LR statistic
    LR_obs <- -2 * (logL0 - logL_hat)
    
    # Bootstrap 
    LR_star <- numeric(B)
    
    for (b in 1:B) {
        # Generate under H0
        bs_sample <- rexp(n, rate = 1/lambda0_hat)
        
        # Refit Weibull
        fit_alt_b <- optim(
            par = c(beta_start, lambda_start),
            fn = function(p) -loglik_weibull(p, bs_sample),
            method = "L-BFGS-B",
            lower = c(1e-6, 1e-6)
        )
        
        logL_star <- -fit_alt_b$value
        
        # Refit exponential
        lambda0_star <- mean(bs_sample)
        logL_star0 <- loglik_exp(lambda0_star, bs_sample)
        
        # LR statistic
        LR_star[b] <- -2 * (logL_star0 - logL_star)
    }
    
    # Bootstrap p-value
    p_value <- (sum(LR_star >= LR_obs) + 1) / (B + 1)
    
    return(list(
        LR_obs = LR_obs,
        p_value = p_value,
        beta_hat = beta_hat,
        lambda_hat = lambda_hat,
        lambda0_hat = lambda0_hat
    ))
}

result <- bootstrap_lrt_weibull(x)
print(result$p_value)
    
```
# Part e

The two tests generated the same results - namely, that the recommended model for the data is a Weibull model, whose scale parameter is not fixed at 1 (and is therefore not exponentially distributed data), and whose maximum likelihood estimate for scale parameter is 31.42 and whose maximum likelihood estimate for shape parameter is 3.37.

